\documentclass[twoside]{article}
\usepackage{stex,physml}
\usepackage[today,fancyhdr]{svninfo}
\usepackage[show]{ed}

\title{The Content Dictionary for SI Dimensions}
\author{Michael Kohlhase}

\begin{document}
\svnInfo $Id: dimensions.tex 44 2008-02-11 16:41:23Z kohlhase $
\svnKeyword $URL: https://svn.omdoc.org/repos/physml/cds/base/dimensions.tex $
\maketitle

\begin{module}[id=dimensions]
\importmodule[../math/arith]{arith}

\symdef{mass}{\mbox{\sf{mass}}}
\begin{consymb}[name=mass]
  The mass dimension.
\end{consymb}

\symdef{length}{\mbox{\sf{length}}}
\begin{consymb}[name=length]
  The length dimension.
\end{consymb}

\symdef{time}{\mbox{\sf{time}}}
\begin{consymb}[name=time]
  The time dimension.
\end{consymb}

\symdef{charge}{\mbox{\sf{charge}}}
\begin{consymb}[name=charge]
  The charge dimension.
\end{consymb} 

\symdef{temperature}{\mbox{\sf{temperature}}}
\begin{consymb}[name=temperature]
  The temparature dimension.
\end{consymb} 

\symdef{intensity}{\mbox{\sf{intensity}}}
\begin{consymb}[name=intensity]
  The intensity dimension.
\end{consymb}

\symdef{angle}{\mbox{\sf{angle}}}
\begin{consymb}[name=angle]
  The angle dimension.
\end{consymb}

\symdef{volume}{\mbox{\sf{volume}}}
\begin{consymb}[name=volume]
  The volume.
\end{consymb}
\begin{definition}[for=volume,type=simple]
  The volume dimension is $\power\length3$
\end{definition}

\symdef{force}{\mbox{\sf{force}}}
\begin{consymb}[name=force]
  The force dimension.
\end{consymb}
\begin{definition}[for=force,type=implicit]
  $\force=\fraction{\times{\mass,\length}}{\power\time2}$.
\end{definition}

\symdef{energy}{\mbox{\sf{energy}}}
\begin{consymb}[name=energy]
  The energy dimension.
\end{consymb}
\begin{definition}[for=energy,type=implicit]
  $\energy=\times{\force,\length}$.
\end{definition}

\end{module}
\end{document}
